---
title: "P8105 Homework 2"
author: "Madison Goldrich - mpg2166"
output: github_document
---

```{r}
library(tidyverse)
```


## Problem 1

First, clean the data in pols-month.csv

```{r}
pols_month_df =
  read_csv("data/pols-month.csv") |> 
  separate(
    mon,
    sep = "-",
    into = c("year", "month", "day")
  ) |> 
  mutate(
    month = case_match(
      month,
      "01" ~ "Jan",
      "02" ~ "Feb",
      "03" ~ "Mar",
      "04" ~ "Apr",
      "05" ~ "May",
      "06" ~ "Jun",
      "07" ~ "Jul",
      "08" ~ "Aug",
      "09" ~ "Sep",
      "10" ~ "Oct",
      "11" ~ "Nov",
      "12" ~ "Dec"
    )
  ) |> 
  mutate(
    president = case_match(
      prez_gop,
      0 ~ "dem",
      1 ~ "gop"
     )
    ) |> 
  select(-day, -prez_gop, -prez_dem)
```

Second, clean the data in snp.csv

```{r}
snp_df =
  read_csv("data/snp.csv") |> 
    separate(
    date,
    sep = "/",
    into = c("month", "day", "year")
  ) |> 
  mutate(
    year = as.numeric(year),
    month = as.numeric(month)
  ) |> 
  mutate(
    year = case_when(
      year <=15 ~ year + 2000,
      year > 15 ~ year + 1900
    )
  ) |> 
  select(year, month, close) |> 
  arrange(year, month) |> 
  mutate(
    month = case_match(
      month,
      1  ~ "Jan",
      2  ~ "Feb",
      3  ~ "Mar",
      4  ~ "Apr",
      5  ~ "May",
      6  ~ "Jun",
      7  ~ "Jul",
      8  ~ "Aug",
      9  ~ "Sep",
      10 ~ "Oct",
      11 ~ "Nov",
      12 ~ "Dec"
    )
  ) |> 
  mutate(
    year = as.character(year)
  )
```

Third, tidy the unemployment data so that it can be merged with the previous datasets.

```{r}
unemployment_df =
  read.csv("data/unemployment.csv") |> 
  pivot_longer(
    Jan:Dec,
    names_to = "month",
    values_to = "unemployment"
  ) |> 
  janitor::clean_names() |> 
  mutate(
    year = as.character(year)
  )
```

Join the datasets

```{r}
pols_snp_df =
  left_join(pols_month_df, snp_df)

five_thirty_eight_df =
  left_join(pols_snp_df, unemployment_df)
```

The pols-month dataset initially contained 9 variables related to the number of national politicians (i.e., president, senators, representatives, etc.) who were democratic or republican at any given time. The snp dataset contained the closing values of the S&P stock index (`close`) on the associated date. Unemployment dataset contained the percentage of unemployment (`unemployment`) in the associated month/year. The resulting merged dataset is `r nrow(five_thirty_eight_df)` rows by `r ncol(five_thirty_eight_df)` columns. Years in which data was collected range from `r min(pull(five_thirty_eight_df, year))` to `r max(pull(five_thirty_eight_df, year))`.


## Problem 2

```{r}
library(readxl)
```

Read and clean the Mr. Trash Wheel sheet.

```{r}
mr_trash_wheel_df =
  read_excel(
    "data/202207 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    range = "A2:N549") |> 
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  separate(
    date,
    sep = "-",
    into = c("year_old", "month_old", "day")
  ) |> 
  select(
    -year_old, -month_old
  ) |> 
  mutate(
    homes_powered,
    homes_powered = (weight_tons * 500) / 30,
    trash_wheel = "Mr.",
    year = as.numeric(year)
  ) |> 
  select(
    trash_wheel, month, day, year, everything()
  )
```

Import and clean Professor Trash Wheel and Gwynnda.

```{r}
prof_trash_wheel_df =
  read_excel(
    "data/202207 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    range = "A2:M96") |> 
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  separate(
    date,
    sep = "-",
    into = c("year_old", "month_old", "day")
  ) |> 
  select(
    -year_old, -month_old
  ) |> 
  mutate(
    homes_powered,
    homes_powered = (weight_tons * 500) / 30,
    trash_wheel = "Professor"
  ) |> 
  select(
    trash_wheel, month, day, year, everything()
  )

gwynnda_trash_wheel_df =
  read_excel(
    "data/202207 Trash Wheel Collection Data.xlsx",
    sheet = "Gwynnda Trash Wheel",
    range = "A2:K108") |> 
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  separate(
    date,
    sep = "-",
    into = c("year_old", "month_old", "day")
  ) |> 
  select(
    -year_old, -month_old
  ) |> 
  mutate(
    homes_powered,
    homes_powered = (weight_tons * 500) / 30,
    trash_wheel = "Gwynnda"
  ) |> 
  select(
    trash_wheel, month, day, year, everything()
  )
```

Combine all 3 trash wheel datasets.

```{r}
mr_prof_df =
  full_join(mr_trash_wheel_df, prof_trash_wheel_df)

trash_wheels_df =
  full_join(mr_prof_df, gwynnda_trash_wheel_df)
```


Each of the individual trash wheel datasets (Mr., Professor, and Gwynnda) contained information about trash collected by these 3 trash wheels. The data was collected on the dates specified and measured things such as the weight of trash collected in tons `weight_tons`, an estimate of the number of homes powered calculated based on the weight of trash collected `homes_powered`, and the amount of specific trash items that were collected - i.e., `plastic_bottles` or `cigarette_butts`. Not all 3 trash wheels measured the exact same specific trash items. Each row in the dataset is an observation, so there are `r nrow(trash_wheels_df)` observations in the `trash_wheels_df` dataset. The total weight of trash collected by Professor Trash Wheel was `r sum(pull(prof_trash_wheel_df, weight_tons))` tons. Gwynnda collected `r sum(pull(filter(gwynnda_trash_wheel_df, month == "July" & year == "2021"), cigarette_butts))` cigarette butts in July 2021.


## Problem 3

Import, clean, and tidy the dataset of baseline demographics.

```{r}
mci_bl_df =
  read_csv("data/MCI_baseline.csv", skip = 1) |> 
  janitor::clean_names() |> 
  mutate(
    age_at_onset = replace(
      age_at_onset, age_at_onset == ".", NA
    ),
    sex = case_match(
      sex,
      1 ~ "male",
      0 ~ "female"
    ),
    apoe4 = case_match(
      apoe4,
      1 ~ "carrier",
      0 ~ "non-carrier"
    )
  ) |> 
  filter(
    is.na(age_at_onset) |
    (as.numeric(age_at_onset) - current_age) > 0
  )
```

Discuss important steps in the import process and relevant features of the dataset:

To import, I had to include `skip = 1` because the first row in the csv file included the variable definitions. I used `janitor::clean_names` to clean the variable names. I used `mutate` to replaces dots that indicated missing values in `age_at_onset` with NAs. I also used mutate to make `sex` and `apoe4` non-numeric. To remove participants who did not meet the inclusion criteria, I used `filter` to keep those who either never developed MCI (implying that they did not have MCI at baseline) or those whose onset age was greater than their current age at baseline. The final tidy `mci_bl_df` dataset is `r nrow(mci_bl_df)` rows by `r ncol(mci_bl_df)` columns. `r nrow(mci_bl_df)` participants were recruited. Of these participants, `r sum(!is.na(pull(mci_bl_df, age_at_onset)))` developed MCI. The average baseline age is `r mean(pull(mci_bl_df, current_age))` years.

What proportion of women in the study are APOE4 carriers?

```{r}
bl_fem_carrier_df =
  filter(
    mci_bl_df,
    sex == "female",
    apoe4 == "carrier"
  )

bl_fem_non_df =
  filter(
    mci_bl_df,
    sex == "female",
    apoe4 == "non-carrier"
  )
```

`r nrow(bl_fem_carrier_df) / nrow(bl_fem_non_df) * 100` percent of women in the study are APOE4 carriers.


Import, clean, and tidy the dataset of longitudinally observed biomarker values.

```{r}
mci_amyloid_df =
  read_csv("data/mci_amyloid.csv", skip = 1) |> 
  janitor::clean_names() |> 
  select(
    id = study_id,
    everything()
  )
```

I imported the dataset using a similar process as the baseline demographics one; I needed to include `skip = 1` in the `read_csv` argument for the same reason. I used `janitor::clean_names` again. Then I just used select to rename `study_id` to `id` so that it matches the first dataset. 
